---
title: "Housing units under construct in the US"
author: "Tran Nguyen"
date: "May 31, 2018"
output: html_document
---

#Introduction
The current housing inventory continues to slip (2018), driving up home prices and limiting home buyers across the U.S.
One might wonder whether building a new home is a good investment. Others might wonder what types of housing units they should consider. Analyzing [Housing Units Under Construction data](https://fred.stlouisfed.org/categories/32303) from St Louis Federal Reserve Economic Database (FRED) will give us some insights on how the housing market is going in the next couple of years.


#Data
Data from FRED was available for three main types of housing units as well as the total:

* New Privately-Owned Housing Units Under Construction: 1-Unit Structures  [(UNDCON1USA)](https://fred.stlouisfed.org/series/UNDCON1USA)

* New Privately-Owned Housing Units Under Construction: 2-4 Unit Structures [(UNDCON24USA)](https://fred.stlouisfed.org/series/UNDCON24USA) 

* New Privately-Owned Housing Units Under Construction: 5-Unit Structures or More [(UNDCON5MUNSA)](https://fred.stlouisfed.org/series/UNDCON5MUNSA)

* New Privately-Owned Housing Units Under Construction: Total [(UNDCONTSA)](https://fred.stlouisfed.org/series/UNDCONTSA)

#Analysis
Let's start by loading necessary packages

```{r}
library(tidyverse)
library(quantmod)
library(data.table)
library(viridis)
library(htmlTable)
library(knitr)
```

Load available data from the website on to the dataframe _my.names_

```{r}
my.names <- data.table(var=c("UNDCON1USA","UNDCON24USA","UNDCON5MUNSA","UNDCONTSA"),
                       name=c("1-unit","2-4 unit","5-unit","total"),
                       Description=c("1-Unit Structures",
                                     " 2-4 Unit Structures",
                                     "5-Unit Structures",
                                     "Total"),
                       Source=c("U.S. Bureau of the Census",
                                "U.S. Bureau of the Census",
                                "U.S. Bureau of the Census",
                                "U.S. Bureau of the Census"))
```

Make a descriptive data table

```{r}
htmlTable(my.names, caption="Data description")
```

Then, use Quantmod to load data

```{r}
df= getSymbols('UNDCON1USA',src='FRED', auto.assign=F) 
```

Next, we slice off necessary data and assign to appropriate smaller dataframes

```{r}
df = data.frame(date=time(df), coredata(df) )

df.24 =getSymbols('UNDCON24USA',src='FRED', auto.assign=F) 
df.24 = data.frame(date=time(df.24), coredata(df.24) )

df.5=getSymbols('UNDCON5MUNSA',src='FRED', auto.assign=F) 
df.5 = data.frame(date=time(df.5), coredata(df.5) )

df.all= getSymbols('UNDCONTSA',src='FRED', auto.assign=F) 
df.all = data.frame(date=time(df.all), coredata(df.all) )
```

Merge and consolidate all data by "date" to graph them together in 1 graph

```{r}
#Merge
df3<-merge(df,df.24,by="date")
df3<-merge(df3,df.5,by="date")
df3<-merge(df3,df.all,by="date")
dt<-data.table(df3)
# Consolidate the data 
dt %>% gather(var,value,-date) %>% data.table() ->dt2

# Merge on variable names
dt2<-merge(dt2,my.names,by="var")
```

Transform data prior to plotting to make the data look 'normal'
```{r}
dt2=dt2[,id:=1:.N, by=var]  # Index running from 1:N by group (var)
dt2=dt2[,var0:=100*value/sum(ifelse(id==1,value,0)),by=var] #create index
```

Finally, we can plot and have a good look at the data

```{r}
# Create caption
mycaption<- "Source: U.S. Bureau of the Census. All are seasonally adjusted."

# Wrap caption 120 characters:
mycaption <- paste0(strwrap(mycaption, 120), sep="", collapse="\n")


# Create Plot
ggplot(data=dt2,aes(x=date,y=var0,color=name,linetype=name))+
  geom_line(size=1.1)+
  scale_y_log10(breaks=c(25,50,100,200,400))+
  theme_minimal()+theme(plot.caption=element_text(hjust=0),
                        legend.position="top")+
  guides(linetype=F)+
  scale_color_viridis(name="Variable",discrete=T,end=0.9)+
  labs(x="",y="Index, January 1970= 100 (log scale)",
       title=" US New Privately-Owned Housing Units Under Construction",
       caption=mycaption)

```



# Result

While new construction for duplexes (2-units) has been flat since the Great Depression of 2008, construction of other types of units have been increasing since then. This translates to an increase in total new construction and can potentially quench the thirst for the housing market in the near future.
Are Americans no longer interested in duplexes? 











